Guarded Recursion in Agda via Sized Types

### 2019-04-18 code-oneclock-paper/latex/Introduction.tex
url
started camera ready

Utilizing techniques for representing ``type theory in type theory''
\cite{AltenkirchK16,Chapman09}, we present an Agda formalization of
the syntax of a simple type theory for guarded recursion. This object
language, which we call \GTT, is a variant of Atkey and McBride's type
system for productive coprogramming \cite{atkey2013productive} in
which the clock context can contain at most one clock. The types of
\GTT\ include the aforementioned \IC{▻} modality and a \IC{□} modality,
a nameless analogue of Atkey and McBride's universal clock
quantification. Clouston \etal \cite{CloustonBGB15} studied a
guarded variant of lambda calculus extended with a \IC{□} operation,
which they call ``constant''.  Our object calculus differs from
theirs in that our judgments are indexed by a clock context, which can
be empty or containing exactly one clock. 
%In this respect \GTT\ is closer in spirit to Atkey and McBride's
%language.
The design of \GTT\ has the benefit of allowing a more appealing introduction rule for
\IC{□} than Clouston et al.'s.

@@

@@

This paper only include the essential parts of our Agda
formalization. The full code is available at
\url{https://github.com/niccoloveltri/agda-clott}. The formalization
uses Agda 2.5.4.2 and Agda standard library 0.16. The paper is
extracted from a literate Agda file, which implies that all the
displayed code has passed Agda's type and termination checker.

>>>>

Utilizing techniques for representing ``type theory in type theory''
\cite{AltenkirchK16,Chapman09}, we present an Agda formalization of
the syntax of a simple type theory for guarded recursion. This object
language, which we call \GTT, is a variant of Atkey and McBride's type
system for productive coprogramming \cite{atkey2013productive} in
which the clock context can contain at most one clock. The types of
\GTT\ include the aforementioned \IC{▻} modality and a \IC{□} modality,
a nameless analogue of Atkey and McBride's universal clock
quantification. Clouston \etal\ \cite{CloustonBGB15} studied a
guarded variant of lambda calculus extended with a \IC{□} operation,
which they call ``constant''.  Our object calculus differs from
theirs in that our judgments are indexed by a clock context, which can
be empty or containing exactly one clock. 
%In this respect \GTT\ is closer in spirit to Atkey and McBride's
%language.
The design of \GTT\ has the benefit of allowing a more appealing introduction rule for
\IC{□} than Clouston et al.'s.

@@

The decision to restrict the clock context of \GTT\ to contain at 
most one clock variable was dictated by 
our choice to use Agda with sized types as the metatheory. In Agda, 
it is possible to encode semantic clock contexts containing multiple 
clocks as lists of sizes or as functions from a finite interval of 
natural numbers into sizes. But Agda's support for sized types is 
tailored to types depending on exactly one size, or on a finite but 
precise number of sizes, which makes it cumbersome to work with types 
depending on clock contexts containing an indefinite number of clocks. 
More technically, as it was privately communicated to us by Agda 
implementors, Agda's type inference only works properly for 
first-order size contraints. We believe that if future Agda 
releases will add the capability of handling multiple sizes, then it would 
be possible to extend the semantics of \GTT\ to a model for the full 
Atkey and McBride's type theory. We also believe that extending our 
formalization to the multiclock case, while technically challenging, 
will not be conceptually harder. 
\GTT\ is stricly less expressive than Atkey and McBride's calculus, since we are 
unable to implement nested coinductive types. But the one clock 
variant of the calculus still allows the construction of a large 
class of coinductive types and it is the subject of active research
\cite{CloustonBGB15}.

@@

This paper only include the essential parts of our Agda
formalization. The full code is available at
\url{https://github.com/niccoloveltri/agda-gtt}. The formalization
uses Agda 2.5.4.2 and Agda standard library 0.16. The paper is
extracted from a literate Agda file, which implies that all the
displayed code has passed Agda's type and termination checker.

### 2019-04-18 code-oneclock-paper/latex/paper.tex
started camera ready

\acknowledgements{We are thankful to Andreas Abel, Guillaume Allais, Herman Geuvers,
  Rasmus Ejlers M{\o}gelberg and Andrea Vezzosi for discussions and
valuable hints.}%optional

%\nolinenumbers %uncomment to disable line numbering

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}

@@

\subsection
{Guarded Recursive Types}
\label{sec:grt}
\input{CloTT/TypeFormers/Mu}

\section{Interpretation and Consistency}
\label{sec:consistent}
\input{Prelude/Interpretation}
\input{CloTT/InterpretSyntax}
\input{GuardedTT}

@@

>>>>

\acknowledgements{We are thankful to Andreas Abel, Guillaume Allais, Herman Geuvers,
  Rasmus Ejlers M{\o}gelberg and Andrea Vezzosi for discussions and
valuable hints. We thank the anonymous referees for their useful comments.}%optional

\nolinenumbers %uncomment to disable line numbering

@@

%Editor-only macros:: begin (do not touch as
%author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{Herman Geuvers}
\EventNoEds{1}
\EventLongTitle{4th International Conference on Formal Structures for Computation and Deduction (FSCD 2019)}
\EventShortTitle{FSCD 2019}
\EventAcronym{FSCD}
\EventYear{2019}
\EventDate{June 24--30, 2019}
\EventLocation{Dortmund, Germany}
\EventLogo{}
\SeriesVolume{131}
\ArticleNo{33}

@@

\subsection{Guarded Recursive Types}
\label{sec:grt}
\input{CloTT/TypeFormers/Mu}

\section{Soundness and Consistency}
\label{sec:consistent}
\input{Prelude/Interpretation}
\input{CloTT/InterpretSyntax}
\input{GuardedTT}

@@

In \GTT\, we restrict the least fixpoint 
operator \IC{μ} to act exclusively on strictly positive functors. This restriction is already 
present in Atkey and McBride's calculus, which is aimed at 
encoding coinductive types using the later modality and it does 
not allow solving general guarded recursive domain equations.
% such as\Ar{D} = ▻ \Ar{D → D}.
\GTT\ is a variant of Atkey and McBride's type theory, therefore the
similar restriction to stricly positive functors.  From the
formalization perspective, this restriction allows us to use Agda's
inductive types to model the guarded recursive types of GTT, as shown
in Section \ref{sec:grt}. In future work, we plan to extend \GTT\ with
the possibility of solving general guarded recursive domain
equations. This could be done in two ways: extending the language with
a universe, which allows the encoding of guarded recursive types as
fixpoints in the universe as shown e.g. by M{\o}gelberg
\cite{Mogelberg14}; or considering a \IC{μ} type former which, besides
stricly positive functors, also operates on functors where all
variables are guarded by an occurrence of the later modality. In the
second option we have to consider strictly positive functors to encode
usual inductive types, such as the natural numbers.
